---
title: "MiCM: visualization"
author: "Ariel Madrigal Aguirre"
date: "`r Sys.Date()`"
output:
  html_notebook:
    df_print: paged
    code_folding: show
    toc: yes
    toc_float: 
      collapsed: false
      smooth_scroll: false
---

```{r,echo=FALSE}
htmltools::img(src = knitr::image_uri("images/micm_color_logo.png"), 
               alt = 'logo', 
               style = 'position:absolute; top:30px; right:0; padding:10px; max-width:50%;')
```

This workshop will guide you through the basics of ChIP-seq analysis with hands-on exercises. You will learn how to process ChIP-seq data: perform read alignment, peak calling, quality control, visualization through the genome browser, motif finding and gene set enrichment analysis.

In this tutorial, we are going to perform motif finding and gene set enrichment analysis. For both tasks, we are going to use a bed file from the data folder, which has the peak calls for HIF1a across the whole genome.

**Note**: All the code in this tutorial will be performed in the terminal. 

Let's move to the scripts directory (or stay here if already there):
```{bash, eval=F}
cd chip_seq/scripts
```
Let's take a look at our bed file:
```{bash}
head ../data/complete/RCC4_Normoxia_HIF1a_PM14_Rep1.bed
```
Count the number of peaks:
```{bash}
wc -l ../data/complete/RCC4_Normoxia_HIF1a_PM14_Rep1.bed
```
# Motif finding
For motif finding, we are going to use [MEME](https://meme-suite.org/meme/tools/meme-chip). For this tutorial, we are going to use the online interactive suite, but you can download/install meme locally by following the instructions in [here](https://meme-suite.org/meme/doc/download.html). There is also a R package that connects the installation to R, you can check it [here](https://bioconductor.org/packages/release/bioc/html/memes.html).  

If the online job takes too long, you can also explore the results in here:
```{bash}
ls ../answers/meme/
```
Open the meme chip [results](../answers/meme/meme-chip.html)

# Gene set enrichment analysis
For gene set enrichment analysis, we are going to use [GREAT](http://great.stanford.edu/public/html/index.php). For this tutorial, we are going to use the online interactive suite, but you can download/install GREAT locally by following the instructions in [here](https://great-help.atlassian.net/wiki/spaces/GREAT/pages/655416/Download). There is also a R package that allows you to use GREAT, you can check it [here](https://bioconductor.org/packages/release/bioc/html/rGREAT.html).


```{r}
sessionInfo()
```

